---
output:
  html_document: default
  pdf_document: default
---

<!-- # : Lane_County_Jail.rmd -->

<!-- # Author: @russl_corey -->

<!-- # Date: Mar 7, 2023 -->

<!-- # This program is free software: you can redistribute it and/or modify it under  -->

<!-- # the terms of the GNU General Public License as published by the Free Software  -->

<!-- # Foundation, either version 3 of the License, or (at your option) any later  -->

<!-- # version. -->

<!-- #  -->

<!-- # This program is distributed in the hope that it will be useful, but WITHOUT ANY  -->

<!-- # WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A  -->

<!-- # PARTICULAR PURPOSE. See the GNU General Public License for more details. -->

<!-- # You should have received a copy of the GNU General Public License along with  -->

<!-- # this program. If not, see <https://www.gnu.org/licenses/>.  -->

---
title: "Lane County Jail Analysis"
author: Russell_Corey
date: March 7, 2023
output: html_document
---

\newpage

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(stringr)
library(tufte)
```



```{r set working director and load data, echo=FALSE, message=FALSE, warning=FALSE}

setwd('/home/russell/Dropbox/DataAnalysis/Lane_County_Jail')

# Use the date that we downloaded booking data
study_date <-  as.Date("2023-04-13")

# define folder and file names based on study date
csv_folder <- paste0("/home/russell/Documents/scrape_jail/",study_date, '/')
charges_file <- paste0('data/charges_', study_date, '.csv')

if (file.exists(charges_file)){
  print(paste(charges_file, 'already exists.'))
} else {
  # Make a list of all the available csv files
  files <- paste0(csv_folder, list.files(csv_folder, pattern='.csv$'))
  
  # init empty var for data
  csv_data <- c()
  
  for(file in files){
    # read csv file
    data <- read_csv(file)
    
    # append loaded data to data frame
    csv_data <- rbind(csv_data, data)
    
    # clean
    rm(data)
  }
  
  # save records
  write_csv(csv_data, charges_file )
  
  rm(csv_folder, csv_data, file, files)
}

```

```{r load charges from csv file, warning=FALSE, echo=FALSE}
# Load charges from csv file
charges <- read_csv(charges_file, show_col_types = FALSE)

# Parse the code from the description
charges$v_code <- gsub("^(\\d+\\.?\\d*).*", "\\1", charges$`Violation:`)

# Get booking information from charges
inmates <- charges %>%
  select(`Booking Number:`, `full name`, `Booking Date:`, `Sched. Release:`, 
         `Released:`, `Sex:`, `Age:`, `Race:`) %>%
  distinct()

# Format dates
inmates$`Sched. Release:` <- as.Date(inmates$`Sched. Release:`, format = "%m/%d/%Y")
inmates$`Released:` <- as.Date(inmates$`Released:`, format = "%m/%d/%Y")

# drop no name charges
charges <- charges[!is.na(charges$`full name`),]

# drop cleared charges
charges <- charges[is.na(charges$`Clearance:`),]
```

```{r label charges based on ORS codes, echo=FALSE, warning=FALSE}
# Load the relevant ORS codes
file <- readLines("/home/russell/Dropbox/DataAnalysis/Lane_County_Jail/data/felonies_and_violent_offenses.txt")
measure_11 <- read.csv("/home/russell/Dropbox/DataAnalysis/Lane_County_Jail/data/codes_measure_11.csv", sep=',')
violent_felonies <- read.csv("/home/russell/Dropbox/DataAnalysis/Lane_County_Jail/data/codes_violent_felonies.csv", sep=':')

# Label the charges 
charges$measure_11 <- ifelse(charges$v_code %in% measure_11$code, TRUE, FALSE)
charges$violent_felonies <- ifelse(charges$v_code %in% violent_felonies$code, TRUE, FALSE)
charges$felony <- ifelse(charges$v_code %in% file, TRUE, FALSE)

# cleanup
rm(measure_11, class_a_felonies, violent_felonies, file)
```

```{r label inmates, warning=FALSE, echo=FALSE}

# filter for measure 11 charges
charges_m11 <- charges[charges$measure_11 == TRUE,]

# filter for violent felonies
charges_vo <- charges[charges$violent_felonies == TRUE,]

# filter for any felony
charges_felony <- charges[charges$felony == TRUE,]


# Filter for only m11 charges then get distinct booking info
m11_booking <- charges_m11 %>%
  select(`Booking Number:`, `full name`, `Booking Date:`, `Sched. Release:`, 
         `Released:`, `Sex:`, `Age:`, `Race:`) %>%
  distinct()

# Filter for only violent felony charges then get distinct booking info
violent_felony_booking <- charges_vo %>%
  select(`Booking Number:`, `full name`, `Booking Date:`, `Sched. Release:`, 
         `Released:`, `Sex:`, `Age:`, `Race:`) %>%
  distinct()

# Filter for only felony charges then get distinct booking info
felony_booking <- charges_felony %>%
  select(`Booking Number:`, `full name`, `Booking Date:`, `Sched. Release:`, 
         `Released:`, `Sex:`, `Age:`, `Race:`) %>%
  distinct()
```


# Intro
This report aims to provide insight into the current detained population in the 
Lane County Jail on `r format(study_date, '%B %d, %Y')`. We will use a Python web scrapper to download the data from the 
jail's online inmate search, then format that data into a single table that we 
will then inspect and generate statistics about the population therein.

Source at [https://github.com/russl-corey/Lane_County_Jail](https://github.com/russl-corey/Lane_County_Jail)

# Data

### Lane County Jail 
Inmate data was retrieved from the Lane County Jail's booking search web portal:

[http://inmateinformation.lanecounty.org/Home/BookingSearchQuery](http://inmateinformation.lanecounty.org/Home/BookingSearchQuery)

on `r format(study_date, '%B %d, %Y')`.

### Oregon 2020 Census

The 2020 Oregon census data can be found at 

<https://www.pdx.edu/population-research/census-data-oregon>


\newpage 

# Inmates


## Race


```{r inmate race, echo=FALSE, warning=FALSE}
# load necessary libraries
library(ggplot2)


#  white percentage
white_per <- 88.8
black_per <- 1.3


# Count inmate race
inmate_counts <- inmates %>% 
  group_by(`Race:`) %>%
  summarize(count = n())

# Calc percentage
inmate_counts$percentage <- round(inmate_counts$count / sum(inmate_counts$count) * 100, 1)

# Show table 
inmate_counts %>%
  kable(caption='Inmate Races')

# save white and black inmate percentages as vars
black_inmate_percentage <- inmate_counts$percentage[inmate_counts$`Race:` == 'B'] 
white_inmate_percentage <- inmate_counts$percentage[inmate_counts$`Race:` == 'W'] 

# calculate likelyhood ratios
black_odds <- black_inmate_percentage / black_per
white_odds <- white_inmate_percentage / white_per

black_inmate_odds <- round(black_odds / white_odds, 1)
```

Breakdown of inmates by race. It is probably safe to assume that the 'W' category label
is probably 'White', and similar, 'A' and 'B' likely could be for 'Asian' and 'Black' 
respectively. The other are too obscure to guess at this point.

Also note that as of the 2020 Census, the overall percentage of black people in 
Lane County is 1.3% and 88.8% for white people. While the inmate population is `r black_inmate_percentage`% for black inmates and `r white_inmate_percentage`% for white inmates. Given this disparity, the odds of a black person in Lane County being an inmate of the jail is `r black_inmate_odds` times more likely than a white community member as of `r format(study_date, '%B %d, %Y')`.


```{r inmate race plot, warning=FALSE, echo=FALSE}
# create the pie chart
ggplot(inmate_counts, aes(x="", y=count, fill=`Race:`)) +
  geom_bar(stat="identity", width=1) +
  coord_polar(theta = "y") +
  labs(title = "Inmate Race Ratios", fill = "Race", x = NULL, y = NULL) +
  theme_void() +
  theme(legend.position = c(1.0, 0.8))

```

\newpage

## Bail

An over look of how many inmates are being held without bail and how many have
bail requirements. This shows the proportion of inmates that have cash or bond 
requirements for bail. The people held with bail represent inmates that are being
held because of money.

```{r inmates bail and measure, echo = FALSE, warning=FALSE}

# Parse bail and bond values and add total bail column
bail_amounts <- charges %>%
  mutate(cash = as.integer(gsub("[\\$,]", "", `Req. Cash Amt:`)), 
         bond = as.integer( gsub("[\\$,]", "", `Req. Bond Amt:`))) %>%
  select(`full name`, cash, bond) %>%
  group_by(`full name`) %>%
  summarize(tot_cash = sum(cash), 
            tot_bond = sum(bond),
            total_bail = sum(cash) + sum(bond))

bail_amounts$bail <- 'No Bail'
bail_amounts$bail[bail_amounts$total_bail > 0] <- 'Bail'

# Create formatted table of bail counts
bail_amounts %>%
  group_by(bail) %>%
  summarize(count = n()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  mutate(percentage = paste0(round(percentage, 0), '%')) %>%
  kable(caption="Inmates with/without Bail Requirements")

# Count the number of inmates with each bail type
bail_counts <- table(bail_amounts$bail)

# Create a pie chart to show the proportion of inmates with different bail types
pie(bail_counts, labels = names(bail_counts), col = c('red', "green"))
title(main = "Proportion of Inmates by Bail Type")

```

\newpage

# Charges

## Measure 11

One of the metrics the sheriff has used is concerning inmates awaiting Measure 11 charges. 
Measure 11 defines minimum mandatory sentences for listed felonies. 


> "No pre-trial adults in custody being held on Measure 11 charges have been
> released due to capacity in nearly 10 years (since July 1, 2013)."
> `r tufte::quote_footer('--- Sheriff Cliff Harrold to Board of County Commissioners, on Jan 25, 2023')`

So let us inspect the proportion of inmates that are currently booked on Measure 11 charges.

```{r measure11 stats, echo=FALSE, warning=FALSE}

# Load list of measure 11 codes
measure11 <- read_csv('measure_11.csv', show_col_types = FALSE)

# encode TRUE/FALSE for measure 11 violations
charges$m11 <- FALSE
charges$m11[charges$v_code %in% measure11$code] <- TRUE

# Filter for only m11 charges then get distinct booking info
m11_booking <- charges %>%
  filter(m11 == TRUE) %>%
  select(`Booking Number:`, `full name`, `Booking Date:`, `Sched. Release:`, 
         `Released:`, `Sex:`, `Age:`, `Race:`) %>%
  distinct()

# add m11 column to inmate data frame
inmates$measure11 <- FALSE 
inmates$measure11[inmates$`Booking Number:` %in% m11_booking$`Booking Number:`] <- TRUE

# Count the number of inmates labeled as measure 11 and not
measure11_counts <- table(inmates$measure11)

data.frame(measure11_counts) %>%
  kable(caption="Inmates with Measure 11 Charges")

# Create a pie chart with percentage labels
pie(measure11_counts, 
    labels = paste0(names(measure11_counts), ": ", 
                    round(100*measure11_counts/sum(measure11_counts), 2), "%"), 
    col = c("orange", "lightblue"),
    main = "Proportion of inmates with measure 11 charges")

# Add number annotations
legend("topright", legend = names(measure11_counts), 
       title = "Measure 11", 
       fill = c("orange", "lightblue"))

```

\newpage

## Violent Felonies

The main concern stated by the Sheriff in regards to the continuation of the jail levy 
is the release of violent offenders into the community due to capacity releases. So let's 
inspect the current inmates booked with at least one violent felony offense. These 
offenses include all of the Measure 11 violations because all of the Measure 11 charges are 
violent felonies. Note, the majority of these inmates are also being held pre-trial, 
and thus presumed innocent.


```{r felony stats, echo=FALSE, warning=FALSE}

# add felony column to inmate data frame
inmates$violent_felony <- FALSE 
inmates$violent_felony[inmates$`Booking Number:` %in% violent_felony_booking$`Booking Number:`] <- TRUE

# Count the number of inmates labeled as felon and not
felony_counts <- table(inmates$violent_felony)

data.frame(felony_counts) %>%
  kable(caption="Inmates with Violent Felony Charges")

# Create a pie chart with percentage labels
pie(felony_counts, 
    labels = paste0(names(felony_counts), ": ", 
                    round(100*felony_counts/sum(felony_counts), 2), "%"), 
    col = c("orange", "lightblue"),
    main = "Proportion of inmates with violent felony charges")

# Add number annotations
legend("topright", legend = names(measure11_counts), 
       title = "violent felony", 
       fill = c("orange", "lightblue"))

```



\newpage

## Top Charges

The majority of inmates are booked for multiple charges. This is a breakdown of 
the total volume of charges on the books. Keep in mind that there are many more 
charges than number of inmates. 

```{r violations, echo=FALSE, warning=FALSE}

# Count all Violations
charge_counts <- charges %>%
  group_by(v_code) %>%
  summarize(count = n()) %>%
  arrange(desc(`count`))

# make list of codes to names
vcode_names <- charges %>%
  select(v_code, `Violation:`) %>%
   distinct(v_code, .keep_all = TRUE)

# merge violation names to violation counts
charge_counts <- merge(x=charge_counts, y=vcode_names,
                    by.x='v_code', by.y='v_code',
                    all.x=TRUE, all.y=FALSE)

# sort and display table of top violations
charge_counts %>%
  select(`Violation:`, count) %>%
  arrange(desc(`count`)) %>%
  head(20) %>%
  kable()


```



\newpage


# Appendix A

source code available at <https://github.com/russl-corey/Lane_County_Jail>

This program is free software: you can redistribute it and/or modify it under the terms of the GNU General Public License as published by the Free Software Foundation, either version 3 of the License, or (at your option) any later version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details. You should have received a copy of the GNU General Public License along with this program. If not, see [\<https://www.gnu.org/licenses/\>](https://www.gnu.org/licenses/).
